---
title: "Renv"
author: "Jacobo Dobon"
date: "2026-04-09"
output:
  html_document:
    toc: TRUE
  md_document:
---


[GitHub Link](https://github.com/Jake-Ramirez/All_Coding_Challenges/tree/main/Bonus%20materials)

## RENV annotation

```{r}
#install.packages("renv", repos = "https://rstudio.r-universe.dev") <- better to install on console not in each knit  
#renv::init() <- write directly on the console source pane.
#All the library will be writen and a new folder called "renv" will be created in the project directory.
```

Using .libPaths() allows us to check where R is retrieving packages from.
R will prioritize the project-specific library managed by renv.

The main purpose of renv is to ensure reproducibility of package versions.
Since R packages are frequently updated, running older code can lead to compatibility issues.
With renv, anyone reproducing the analysis can use exactly the same package versions as originally used.

If you want to install a new package within a project managed by renv, you can do it as usual.
However, installing a package does not automatically make it part of the reproducible environment.

Newly installed packages will not appear in the renv.lock file until you explicitly update it using a snapshot.

Example:

```{r}
source("renv/activate.R")
```

If we use the .libPaths() will show us where the package will pull our imputs.Will always try the RPackage first.

The function of renv is for reproducibility of packages versions, because packages changes all the time, so if you are running an project from years ago, there is a possibility that the code will be running through a old version of some packages. Renv will allow the person who are going to reproduce the research to run the code under the same packages versions that was run before.

If you are running into the project under renv and want to install a new package you just do the same code. The difference is that will ask you if you want to install it. However, just because you intalled it, doesn meant that you can use it on the same environment.

The new installed packeges will not be shown into the renv_lock file, because you need to save and ask for renv to update the lockfile using snapshot.

Exemple:

```{r}
#install.packages("paletteer")
```
To save changes, run renv::snapshot() in the console.

When cloning a repository that uses renv, even if package names are listed in the renv.lock file, they are not automatically installed.
To recreate the environment, you must run:

```{r}
renv::restore()
```

---------------------------------------------------------------------------------------------

## In class review annotation 

Subset:

```{r}
mtcars
```

Subsetting

```{r}
mtcars[1,4]

# Find all rows such that VS = 1
mtcars[mtcars$vs == 1,]

mtcars$hp[mtcars$vs == 1]

mtcars$hp[mtcars$vs == 1][1]

mtcars$hp[mtcars$vs == 1 & mtcars$gear == 4]

# Using subset function
subset(mtcars, vs == 1)

subset(mtcars, vs == 1 & gear == 4)

subset(mtcars, vs == 1 | gear == 4)


# Subset with min and max
max(mtcars$hp)
min(mtcars$hp)

```

---------------------------------------------------------------------------------

## Bonus material annotation

[Points on a map](https://github.com/Noel-Lab-Auburn/OomyceteDiversityinAlabamaCotton/blob/main/BiogeographyModeling.R)

### Putting points on a map by Zachary Noel

I decided to explore the topic of mapping data points because of its strong potential for future research applications in my field.
In particular, I am interested in studying transcriptional variation in Homalodisca vitripennis populations in relation to their oral microbiome.

By integrating spatial information with transcriptomic data, it becomes possible to identify geographic patterns associated with specific bacterial communities carried by the insect.
This approach could help reveal how different microbial compositions influence gene expression in the host, and ultimately provide insights into vector-pathogen interactions.

Learning how to visualize data on maps is therefore a key step toward combining spatial, microbiological, and transcriptomic datasets in a meaningful way.

### Class annotation:

Libraries needed

```{r}
#install.packages("tidyverse")
library(tidyverse)
#install.packages("usmap")
library(usmap)
#install.packages("maps")
library(maps)
#install.packages("mapdata")
library(mapdata)
```

Collor palette

```{r}
cbbPalette <- c("black", "#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00")
options(dplyr.print_max = 1e9)

```

Read the data

```{r}
Dairy.mr <- read.csv("Dairy.m.csv")

```

```{r}
counties <- map_data("county")
counties$State < toupper(counties$Region)

```
```{r}
str(Dairy.mr)
milk.production.state <- Dairy.mr %>%
  group_by(State) %>%
  summarise(mean_milk_production = mean(Milk_production_state, na.rm = TRUE))

milk.production.state$State <- toupper(milk.production.state$State)
state.data <- map_data("state")
state.data$State <- toupper(state.data$region)

```

You can also restrict the analysis to specific states:

```{r}
#state.data2 <- state.data %>%
#  subset(State %in% c("ALABAMA",
#                      "FLORIDA",
#                      "TEXAS", 
#                      "ARKANSAS",
#                      "MISSISSIPPI",
#                      "SOUTH CAROLINA",
#                      "NORTH CAROLINA",
#                      "VIRGINIA",
#                      "LOUISIANA",
#                      "TENNESSEE",
#                      "GEORGIA",
#                      "OKLAHOMA"))
```

Example of joining datasets and filtering:

```{r}
# cotton.acres.harvested.bycounty2 <- left_join(counties, cotton.acres.harvested.bycounty, by = c("State", "County"))
# cotton.acres.harvested.bycounty.plot <- cotton.acres.harvested.bycounty2 %>%
#   subset(State %in% c("ALABAMA",
#                       "FLORIDA",
#                       "TEXAS", 
#                       "ARKANSAS",
#                       "MISSISSIPPI",
#                       "SOUTH CAROLINA",
#                       "NORTH CAROLINA",
#                       "VIRGINIA",
#                       "LOUISIANA",
#                       "TENNESSEE",
#                       "GEORGIA",
#                       "OKLAHOMA"))
```

Unit conversion:

```{r}
# cotton.acres.harvested.bycounty.plot$Hectares <- cotton.acres.harvested.bycounty.plot$mean.acres.harvested*0.404686

```

Counting observations per coordinate:

```{r}
# map.points <- seedtreatment.prediction.nona %>%
#   group_by(Long, Lat) %>%
#   tally()
```


I used this example to understand the workflow and keep it as a reference for future wrokflows.


```{r}
milk.by.state.map <- left_join(state.data, milk.production.state, by = "State")

Fig1 <- ggplot() +
  geom_polygon(data = milk.by.state.map, 
               aes(x = long, y = lat, group = group, fill = mean_milk_production),
               color = "grey", size = 0.2) +
  geom_polygon(data = milk.by.state.map, 
               aes(x = long, y = lat, group = group), color = "black", fill = NA) +
  scale_fill_gradient(low = "#ECFFDC", high = "#4F7942", 
                      na.value = "white", 
                      name = "Avg Milk Production (kg)") +
  theme_minimal() +
  coord_map("albers", lat0 = 45.5, lat1 = 29.5) +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Average State-level Milk Production in the US") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  )
Fig1
```

Note: In this dataset, we do not have latitude and longitude for individual observations.
To plot points on the map, we would use:

```{r}
# geom_point(data = map.points, 
#            aes(x = Long, y = Lat, size = n),
#            fill = cbbPalette[[8]], 
#            color = "black", 
#            alpha = 0.6, 
#            shape = 21)
```

Learning how to represent data spatially was very useful, and I expect to apply these techniques in future analyses involving biological and transcriptomic datasets.



