---
title: "Linear Models Challenge VII"
author: "Jacobo Dobon"
date: "2026-04-02"
output: 
  md_document:
    variant: markdown_github
  html_document: default
---

### 1. Read in the data called “PlantEmergence.csv” using a relative file path and load the following libraries. tidyverse, lme4, emmeans, multcomp, and multcompView. Turn the Treatment , DaysAfterPlanting  and Rep into factors using the function as.factor

```{r}
PlantEmergence <- read.csv("PlantEmergence.csv")

library(tidyverse)
library(lme4)
library(emmeans)
library(multcomp)
library(multcompView)

cols <- c("Treatment", "DaysAfterPlanting", "Rep")
PlantEmergence[cols] <- lapply(PlantEmergence[cols], as.factor)
```


### 2. Fit a linear model to predict Emergence using Treatment and DaysAfterPlanting along with the interaction. Provide the summary of the linear model and ANOVA results.

```{r}
lm_Emergence <- lm(Emergence ~ Treatment*DaysAfterPlanting, data = PlantEmergence)
summary(lm_Emergence)
anova(lm_Emergence)
```
### 3. Based on the results of the linear model in question 2, do you need to fit the interaction term? Provide a simplified linear model without the interaction term but still testing both main effects. Provide the summary and ANOVA results. Then, interpret the intercept and the coefficient for Treatment 2.

There is no need because p = 1 so is not significant.

```{r}
lm_simple <- lm(Emergence ~ Treatment + DaysAfterPlanting, data = PlantEmergence)
summary(lm_simple)
anova(lm_simple)
```

The intercept is 182.163, that's the number of plants that emerged at day 7 for treatment 1.
Treatment 2 results in approximately 134.5 fewer emerged plants compared to Treatment 1, holding DaysAfterPlanting constant.


### 4. Calculate the least square means for Treatment using the emmeans package and perform a Tukey separation with the compact letter display using the cld function. Interpret the results.

```{r}
lm_Treatment <- lm(Emergence ~ Treatment, data = PlantEmergence)
emm <- emmeans(lm_Treatment, ~ Treatment)
emm
Results_lsmeans <- cld(emm, Letters = letters, alpha = 0.05, reversed = TRUE, details = TRUE)
Results_lsmeans
```

Treatment 2 forms a separate group, indicating that it has significantly lower emergence compared to the other treatments.


### 5. The provided function lets you dynamically add a linear model plus one factor from that model and plots a bar chart with letters denoting treatment differences. Use this model to generate the plot shown below. Explain the significance of the letters. 

```{r}
plot_cldbars_onefactor <- function(lm_model, factor) {
  data <- lm_model$model
  variables <- colnames(lm_model$model)
  dependent_var <- variables[1]
  independent_var <- variables[2:length(variables)]

  lsmeans <- emmeans(lm_model, as.formula(paste("~", factor))) # estimate lsmeans 
  Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details = TRUE, Letters = letters)
  
# Extracting the letters for the bars
  sig.diff.letters <- data.frame(Results_lsmeans$emmeans[,1], 
                                 str_trim(Results_lsmeans$emmeans[,7]))
  colnames(sig.diff.letters) <- c(factor, "Letters")
  
  # for plotting with letters from significance test
  ave_stand2 <- lm_model$model %>%
    group_by(!!sym(factor)) %>%
    dplyr::summarize(
      ave.emerge = mean(.data[[dependent_var]], na.rm = TRUE),
      se = sd(.data[[dependent_var]]) / sqrt(n())
    ) %>%
    left_join(sig.diff.letters, by = factor) %>%
    mutate(letter_position = ave.emerge + 10 * se)
  
  plot <- ggplot(data, aes(x = !! sym(factor), y = !! sym(dependent_var))) + 
    stat_summary(fun = mean, geom = "bar") +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
    ylab("Number of emerged plants") + 
    geom_jitter(width = 0.02, alpha = 0.5) +
    geom_text(data = ave_stand2, aes(label = Letters, y = letter_position), size = 5) +
    xlab(as.character(factor)) +
    theme_classic()
  
  return(plot)
}

plot_cldbars_onefactor(lm_Treatment, "Treatment")
```

Same interpretation as in the previous answer, treatment 2 significantly reduces plant emergence up to 1/4 compared to the rest of the treatments.





