---
title: "Loops and Functions"
author: "Jacobo"
date: "2026-03-19"
output: html_document
---

First function we do is converting ºF to ºC to avoid copy pasting or defining a new object each time
```{r}
F_to_C <- function(fahrenheit_temp){
  celsius <- (5*(fahrenheit_temp - 32)/9)
  return(celsius)
}
#we call the function
F_to_C(32)
```

Function to convert back ºC to ºF
```{r}
C_to_F <- function(celsius_temp){
  fahrenheit <- celsius_temp*9/5+32
  return(fahrenheit)
}

#and call the function
C_to_F(0)
```

Testing the combination of rep and seq
```{r}
rep(seq(from = 0, to = 10, by = 2), times = 3, each = 2)
```

A simple for loop:
```{r}
for (i in 1:10) {
print(i*2)  
}
```

Now a for loop that uses a function we previously created:
```{r}
for (i in -30:100){
  result <- F_to_C(i)
  print(result)
}
```

But now we want to actually save somewhere the outcome so we create a Null object that we will be appending the outcome of the iteration each time:
```{r}
celcius.df <- NULL 
for (i in -30:100){   
result_i <- data.frame(F_to_C(i), i)   
celcius.df <- rbind.data.frame(celcius.df, result_i) 
}
head(celcius.df)
```






